Timing and magnitude of climate driven range shifts in transboundary fish stocks challenge their management

### 2021-11-15 Scripts/Initial_Analysis.Rmd
First draft towards co-authors

@@

  # ToE Map 
  toe_map_data <- World_eez_sf %>% 
    left_join(Eez_d_hist,
              by = "eezid") %>% 
    filter(tresh == tresh_list[i],
           metric == "mean_yr")

@@

@@

##### Fig. N3 Mapping the time of emergence Of Top stocks

@@

World_map_top <- World_eez_sf %>% 
  left_join(Matching_names) %>% 
  full_join(Top_map_data,
            by = c("sau_fishing_entity")
  ) %>% 
  filter(!is.na(variable))

@@

  no_toe_eez <- World_eez_sf %>% 
    filter(eez_name %in% Neighbours$eez_name, # Filter EEZs without Transboundary spp
           !eez_name %in% World_map_top$eez_name) %>%  # Filter EEZs with ToE stocks
    rbind(russia_churchil)

  no_toe_land <- World_map %>% 
    filter(name %in% Neighbours$eez_name, # Filter EEZs without Transboundary spp
           !name %in% World_map_top$name) # Filter EEZs with ToE stocks

  map <- ggplot(World_map_top) +
    geom_sf(
      aes(
        fill = mean #original map
        # fill = value
      ),
      color =  "white",
      size = 0.1
    ) +
    # grey maps
    # ggplot() +
    geom_sf(data = World_map, aes(), fill = "grey50", color =  "white",size = 0.1) +
    geom_sf(data = no_toe_eez, aes(), fill = "lightblue", color =  "white",size = 0.1) +
    # geom_sf(data = no_toe_land, aes(), fill = "lightblue", color =  "white",size = 0.1) +
    # Set Participation labels
    scale_fill_viridis("Average year of range shift of top valued transboundary stocks",
                       limits = c(2000,2100),
                       breaks = seq(2000,2100,25),
                       direction = -1
    ) +
    my_ggtheme_m(leg_pos = "bottom") +
    theme(legend.key.width = unit(5, "line")) +
    guides(
      fill=guide_colorbar(title.position="top", 
                        title.hjust =0.5)
  )

  ggsave(
    plot = map,
    "./Figures/top_toe_map.png",
    width = 10,
    height = 5,
    units = "in"
  )

@@

# Convert to data to spatial
top_network_data_sf <- top_data %>%
  select(start_lon, start_lat, end_lon, end_lat) %>% 
  purrr::transpose() %>% 
  purrr::map(~ matrix(flatten_dbl(.), nrow = 2, byrow = TRUE)) %>% 
  purrr::map(st_linestring) %>%
  st_sfc(crs = 4326) %>%
  # st_sfc(crs = "+proj=eck4") %>% # for tbular plot
  st_sf(geometry = .)%>%
  bind_cols(top_data) %>%
  select(everything(), geometry) %>% 
  # Set circles ('cus the world is not flat!)
  st_segmentize(units::set_units(100, km)) %>% 
  mutate(geometry = (geometry + c(180,90)) %% c(360) - c(180,90)) %>%
  st_wrap_dateline(options = c("WRAPDATELINE=YES",  "DATELINEOFFSET=180"), quiet = TRUE) %>%
  sf::st_sf(crs = 4326)
  # sf::st_sf(crs = "+proj=eck4") # for tbular plot

@@

>>>>

World_map_nemer <- World_map %>%
  full_join(Value_map_data,
            by = c("sau_fishing_entity","fishing_entity_id")
  ) %>% 
  filter(!is.na(participation)) %>% 
sf::st_transform(crs = "+proj=eck4")

# These are regions without transboundary species
grey_land <- World_map %>% 
  filter(!fishing_entity_id %in% Value_map_data$fishing_entity_id) %>% 
  filter(!sovereignt %in% c("Antarctica"))%>% 
sf::st_transform(crs = "+proj=eck4")

@@

  # ToE Map 
  toe_map_data <- World_eez_sf %>% 
    left_join(Eez_d_hist,
              by = "eezid") %>% 
    filter(tresh == tresh_list[i],
           metric == "mean_yr")%>% 
sf::st_transform(crs = "+proj=eck4")

@@

  russia_churchil <- World_eez_sf %>% 
    filter(eez_name %in% c("Russia (Kara Sea)","Russia (Laptev to Chukchi Sea)"))%>% 
sf::st_transform(crs = "+proj=eck4")

  no_toe_eez <- World_eez_sf %>% 
    filter(eez_name %in% Neighbours$eez_name, # Filter EEZs without Transboundary spp
           !eez_name %in% toe_map_data$eez_name) %>%  # Filter EEZs with ToE stocks
    rbind(russia_churchil)%>% 
sf::st_transform(crs = "+proj=eck4")

  no_toe_land <- World_map %>% 
    filter(name %in% Neighbours$eez_name, # Filter EEZs without Transboundary spp
           !name %in% World_map_part$name)%>% 
sf::st_transform(crs = "+proj=eck4")  # Filter EEZs with ToE stocks

@@

##### Fig. 4N Mapping the time of emergence Of Top stocks

@@

World_map_top <- World_eez_sf %>% 
  left_join(Matching_names) %>% 
  full_join(Top_map_data,
            by = c("sau_fishing_entity")
  ) %>% 
  filter(!is.na(variable)) %>% 
  sf::st_transform(crs = "+proj=eck4")

@@

  russia_churchil <- World_eez_sf %>% 
    filter(eez_name %in% c("Russia (Kara Sea)","Russia (Laptev to Chukchi Sea)"))%>% 
    sf::st_transform(crs = "+proj=eck4")

  no_toe_eez <- World_eez_sf %>% 
    sf::st_transform(crs = "+proj=eck4") %>% 
    filter(eez_name %in% Neighbours$eez_name, # Filter EEZs without Transboundary spp
           !eez_name %in% World_map_top$eez_name) %>%  # Filter EEZs with ToE stocks
    rbind(russia_churchil)%>% 
    sf::st_transform(crs = "+proj=eck4")

  no_toe_land <- World_map %>% 
    filter(name %in% Neighbours$eez_name, # Filter EEZs without Transboundary spp
           !name %in% World_map_top$name
           ) %>% 
    sf::st_transform(crs = "+proj=eck4")

  map <- ggplot(World_map_top) +
    geom_sf(
      aes(
        fill = mean #original map
        # fill = value
      ),
      color =  "white",
      size = 0.1
    ) +
    # grey maps
    # ggplot() +
    geom_sf(data = subset(World_map, sovereignt != "Antarctica"), aes(), fill = "grey50", color =  "white",size = 0.1) +
    geom_sf(data = no_toe_eez, aes(), fill = "lightblue", color =  "white",size = 0.1) +
    # geom_sf(data = no_toe_land, aes(), fill = "lightblue", color =  "white",size = 0.1) +
    # Set Participation labels
    scale_fill_viridis("Average year of range shift of top valued transboundary stocks",
                       limits = c(2000,2100),
                       breaks = seq(2000,2100,25),
                       direction = -1
    ) +
    my_ggtheme_m(leg_pos = "bottom") +
    theme(legend.key.width = unit(5, "line")) +
    guides(
      fill=guide_colorbar(title.position="top", 
                        title.hjust =0.5)
  )

  ggsave(
    plot = map,
    "./Figures/top_toe_map.png",
    width = 8,
    height = 5,
    units = "in"
  )

@@

# Convert to data to spatial
top_network_data_sf <- top_data %>%
  select(start_lon, start_lat, end_lon, end_lat) %>% 
  purrr::transpose() %>% 
  purrr::map(~ matrix(flatten_dbl(.), nrow = 2, byrow = TRUE)) %>% 
  purrr::map(st_linestring) %>%
  st_sfc(crs = 4326) %>%
  # st_sfc(crs = "+proj=eck4") %>% # for tbular plot
  st_sf(geometry = .)%>%
  bind_cols(top_data) %>%
  select(everything(), geometry) %>% 
  # Set circles ('cus the world is not flat!)
  st_segmentize(units::set_units(100, km)) %>% 
  mutate(geometry = (geometry + c(180,90)) %% c(360) - c(180,90)) %>%
  st_wrap_dateline(options = c("WRAPDATELINE=YES",  "DATELINEOFFSET=180"), quiet = TRUE) %>%
  # sf::st_sf(crs = 4326)
  sf::st_sf(crs = "+proj=eck4") # for tbular plot

@@

World_eez_sf <- World_eez_sf %>% 
sf::st_transform(crs = "+proj=eck4")

World_map <- World_map %>% 
sf::st_transform(crs = "+proj=eck4")

### 2021-11-15 Scripts/Manuscript.Rmd
First draft towards co-authors

First, we estimated the catch proportion as the stock share ratio (SSR) catch of each transboundary stock shared by neighboring nations between 1951 and 2100. We did this by aggregating the projected catch within the $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the stock is present across neighboring EEZs, and then calculated the proportion of the stock's catch held within each EEZ [@PalaciosAbrantes:2020tv]. Second, to reduce the effects of variability, we averaged the calculated proportion of stock occupying a given EEZ into three time periods: the historical time period ($SSR_{th}$) from 1951 to 2005, and two future periods, the early $21^{st}$ century, between 2021 and 2040 ($SSR_{te}$), and the mid $21^{st}$ century, defined as the average of 2041 to 2060 ($SSR_{tm}$). These future time periods were chosen to correspond to the challenges of achieving fisheries-relevant UN-SDGs by 2030 (i.e., the 2030 Agenda), such as SDG 14.4 (end overfishing), SDG 2.4 (ensure sustainable food production systems) and SDG 1.2 (poverty reduction) [@Singh:2017ds; @UnitedNations:1Qe8KIcN]. The analysis was replicated for projected stock distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, we defined a threat point for each EEZ’s stock as $SSR_{th} \pm 2 \sigma$, where  is the standard deviation of $SSR_{th}$. Thus, a change of SSR catch beyond an EEZ’s threat point happens when the future SSR catch exceeds two standard deviations of the historical variations in SSR (i.e., when $SSR_{et} \geqq (SSR_{th} + 2 \sigma$) or $SSR_{e,m} \leqq (SSR_{th} - \sigma)$. Finally, we estimated the percentage change in SSR ($\Delta SSR$) of each future time period ($SSR_{tf}$) relative to the historic time period $SSR_{th}$ for each stock whose share ratio exceeded the threat point following [@PalaciosAbrantes:2020tv];

@@

**Most EEZs are expected to see a relative small number of stocks shifting their SSR (Fig. \@ref(fig:toeFigure4)A). On average, EEZs are expected to see changes in SSR on 16% $\pm$ 10 of transboundary stocks increasing to 20% $\pm$ 11 by mid of the $21^{st}$ century. However, this result masks mask large variation in changes across EEZs. By early century, the EEZs of French Guiana (54%), South Africa (59%), Falkland Isl. (61%), Brazil (75%) and Kerguelen Isl., and Pitcairn (both 100%) are projected to see changes in SSR of over 50% in their transboundary stocks. On the other hand, some island territories such as Anguilla are projected to see changes in as low as 2% of their transboundary stocks.**

**Shifts in the SSR of the top 5 most valuable transboundary stocks of each country are projected to be 61 $\pm$ 18% by 2030 and 72 $\pm$ 17% by 2050 in 83% of the world's EEZs (Figure SX). In some cases, the SSR of one EEZ is projected increase in more than double relative to their neighbors (e.g., from 10% to 30% SSR). Some of these EEZs include Guatemala gaining from Mexico in the Pacific, Cyprus from Greece and Turkey in the Mediterranean Sea, and Russia from Norway and Japan in the Barents Sea and Western Pacific, respectively**

@@

Our results identify regional “hotspots” of climate risk for transboundary fisheries management that will require the prompt adaptation of management plans (Figs. \@ref(fig:toeFigure2)A and \@ref(fig:toeFigure4)A). Regions such as the Caribbean are characterized by high levels of warming relative to historical natural variability [@IPCC:2019tn; @Hawkins:2012hc] and species’ high vulnerability to warming waters [@IPCC:2019tn]. Moreover, they represent relatively low area EEZs that border multiple countries. Game theory predicts that the greater the number of negotiating parties, the harder it is for parties to reach an agreement [@Gronbaek:2020ei]. Thus, coordinating the management of shifting transboundary stocks for countries in these regions will be particularly challenging.

@@

Different levels of uncertainty exist in the projected emergence of environmental variables mainly related to *i*) climate change scenarios and *ii*) model structure [@Frolicher:2016dg]. First, our analysis is based on a high emission scenario (RCP 8.5) representing an "extreme" case of shifts in the shared distribution of transboundary species.Thus, any mitigation efforts could result in a substantial delay in distribution shifts as environmental signals are responsive to mitigation, at least after the middle of the $21^{st}$ century [@Frolicher:2016dg]. Second, a substantial source of uncertainty is related to model selection for both climate change and fish distribution. This is specifically important for early time periods (e.g., 2016-2035) where the uncertainty related to model selection, including the parameterization of poorly understood processes that regulate NPP changes, is often larger than the uncertainty stemming from the climate change scenario [@Frolicher:2016dg]. While such uncertainty could potentially be reduced by incorporating ensemble simulations from a range of different ESMs, these are computationally expensive simulations [@Frolicher:2009ge; @Frolicher:2016dg; @Rodgers:2015jq] that are only just becoming available [@Deser:2020bj] and cannot address limitations of process-level understanding. It is unlikely that these complex uncertainties will adequately be addressed in the coming decade. This is critical for the tropics, where the shared distribution of transboundary stocks is expected to shift the earliest, and the response of the base of the food web to climate change is most uncertain [@Tagliabue:2020hd; @Kwiatkowski:2020cf]. These uncertainties also affect the projection of species distributions [@Lotze:2019ce; @BryndumBuchholz:2019kh]. Overall, multiple upper trophic level models present overall agreement in terms of directional change of fish biomass, but variability in the magnitude of that change [@Lotze:2019ce]. Further research that reduces uncertainty in the NPP response at the base of ocean food webs, alongside large ensemble simulations of multiple species distribution models could lead to smaller structural uncertainty of fish and fisheries models [@BryndumBuchholz:2019kh]. **Another important source of uncertainty in this study is the utilization of political boundaries (EEZ) to define stocks, rather than biologically-defined populations. While this method might define some stocks that do not necessarily align with biologically-defined sub-populations within an EEZ, in many EEZs, fisheries are often managed at the species level [@MAP:2017uf] and sub-populations are potentially interconnected [@Ramesh:2019va], thus, providing additional ecological ground for our analysis [@Dunn:2019ca]. Moreover, not limiting the species to, at least, an EEZ could result in an overestimation of our results (NOTE: See Ovando *et al.* https://www.pnas.org/content/118/23/e2025958118).** Reproducing our analysis regionally, where spatially explicit stock data is available would allow to generate better constrained results and potentially identify different types of shared stock shifts at meta-population level [@Link:2010ei; @Archambault:2016ie]. Addressing these uncertainties systematically can serve as a roadmap for future studies to provide additional information to inform policy towards sustainable and equitable international fisheries management under climate change.

>>>>

First, we estimated the catch proportion as the stock share ratio (SSR) catch of each transboundary stock shared by neighboring nations between 1951 and 2100. We did this by aggregating the projected catch within the $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the stock is present across neighboring EEZs, and then calculated the proportion of the stock's catch held within each EEZ [@PalaciosAbrantes:2020tv]. *This way, if, Canada and the US share a specific stock the SSR for Canada could be 20% and that of the US 80%, meaning that Canada holds 20% of the distribution of the stock within its waters and the US the other 80%.*Second, to reduce the effects of variability, we averaged the calculated proportion of stock occupying a given EEZ into three time periods: the historical time period ($SSR_{th}$) from 1951 to 2005, and two future periods, the early $21^{st}$ century, between 2021 and 2040 ($SSR_{te}$), and the mid $21^{st}$ century, defined as the average of 2041 to 2060 ($SSR_{tm}$). These future time periods were chosen to correspond to the challenges of achieving fisheries-relevant UN-SDGs by 2030 (i.e., the 2030 Agenda), such as SDG 14.4 (end overfishing), SDG 2.4 (ensure sustainable food production systems) and SDG 1.2 (poverty reduction) [@Singh:2017ds; @UnitedNations:1Qe8KIcN]. The analysis was replicated for projected stock distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, we defined a threat point for each EEZ’s stock as $SSR_{th} \pm 2 \sigma$, where  is the standard deviation of $SSR_{th}$. Thus, a change of SSR catch beyond an EEZ’s threat point happens when the future SSR catch exceeds two standard deviations of the historical variations in SSR (i.e., when $SSR_{et} \geqq (SSR_{th} + 2 \sigma$) or $SSR_{e,m} \leqq (SSR_{th} - \sigma)$. Finally, we estimated the percentage change in SSR ($\Delta SSR$) of each future time period ($SSR_{tf}$) relative to the historic time period $SSR_{th}$ for each stock whose share ratio exceeded the threat point following [@PalaciosAbrantes:2020tv];

@@

**Most EEZs are expected to see a relatively small number of stocks shifting their SSR (Fig. \@ref(fig:toeFigure4)A). On average, EEZs are expected to see changes in SSR on 16% $\pm$ 10 of transboundary stocks increasing to 20% $\pm$ 11 by mid of the $21^{st}$ century. However, this result masks large variation in changes across EEZs. By early century, the EEZs of French Guiana (54%), South Africa (59%), Falkland Island (61%), Brazil (75%) and Kerguelen Island, and Pitcairn (both 100%) are projected to see changes in SSR of over 50% in their transboundary stocks. On the other hand, some island territories such as Anguilla are projected to see changes in as low as 2% of their transboundary stocks.**

**Shifts in the SSR of the top 5 most valuable transboundary stocks of each country are projected to be 61 $\pm$ 18% by 2030 and 72 $\pm$ 17% by 2050 in 83% of the world's EEZs (Figure SX). In some cases, the SSR of one EEZ is projected to increase more than double relative to their neighbors (e.g., from 10% to 30% SSR). Some of these EEZs include Guatemala gaining from Mexico in the Pacific, Cyprus from Greece and Turkey in the Mediterranean Sea, and Russia from Norway and Japan in the Barents Sea and Western Pacific, respectively**

@@

**The SDG target 17.16 (Enhance the global partnership for sustainable development) calls for collaborations between stakeholders to support the achievement of the SDGs in all countries, in particular developing countries [@SDGREF]. Such collaborations will be critical if we are to sustainably manage transboundary fisheries under climate change [@Pinsky:2018cb; @Oremus:2020id].** Our results identify regional “hotspots” of climate risk for transboundary fisheries management that will require the prompt adaptation of collaborative management plans (Figs. \@ref(fig:toeFigure2)A and \@ref(fig:toeFigure4)A). Regions such as the Caribbean are characterized by high levels of warming relative to historical natural variability [@IPCC:2019tn; @Hawkins:2012hc] and species’ high vulnerability to warming waters [@IPCC:2019tn]. Moreover, they represent relatively low area EEZs that border multiple countries. Game theory predicts that the greater the number of negotiating parties, the harder it is for parties to reach an agreement [@Gronbaek:2020ei]. Thus, coordinating the management of shifting transboundary stocks for countries in these regions will be particularly challenging.

@@

Different levels of uncertainty exist in the projected emergence of environmental variables mainly related to *i*) climate change scenarios and *ii*) model structure [@Frolicher:2016dg]. First, our analysis is based on a high emission scenario (RCP 8.5) representing an "extreme" case of shifts in the shared distribution of transboundary species. Thus, any mitigation efforts could result in a substantial delay in distribution shifts as environmental signals are responsive to mitigation, at least after the middle of the $21^{st}$ century [@Frolicher:2016dg]. Second, a substantial source of uncertainty is related to model selection for both climate change and fish distribution. This is specifically important for early time periods (e.g., 2016-2035) where the uncertainty related to model selection, including the parameterization of poorly understood processes that regulate NPP changes, is often larger than the uncertainty stemming from the climate change scenario [@Frolicher:2016dg]. While such uncertainty could potentially be reduced by incorporating ensemble simulations from a range of different ESMs, these are computationally expensive simulations [@Frolicher:2009ge; @Frolicher:2016dg; @Rodgers:2015jq] that are only just becoming available [@Deser:2020bj] and cannot address limitations of process-level understanding. It is unlikely that these complex uncertainties will adequately be addressed in the coming decade. **NOTE OPTION A. Therefore, decision-makers are facing the challenge of taking action under an uncertain future. One strategy for that is adapting scenario planning into fisheries management [@NOAA_scen_plan_2020]. This will allow for managers to integrate the uncertainty into planning for managing fisheries. For example, if stocks are expected to shift their distribution by 2036 $\pm$ 28 years (Fig. 2), managers could prepare for a scenario where stocks are currently changing and another where they will shift in 20 years. Scenario planning can also be used to estimate potential looses in revenue when adaptation to climate change is not developed cooperatively [@Sumaila:2020tp; @Miller:2013iv].** This is critical for the tropics, where the shared distribution of transboundary stocks is expected to shift the earliest, and the response of the base of the food web to climate change is most uncertain [@Tagliabue:2020hd; @Kwiatkowski:2020cf].

These uncertainties also affect the projection of species distributions [@Lotze:2019ce; @BryndumBuchholz:2019kh]. Overall, multiple upper trophic level models present overall agreement in terms of directional change of fish biomass, but variability in the magnitude of that change [@Lotze:2019ce]. Further research that reduces uncertainty in the NPP response at the base of ocean food webs, alongside large ensemble simulations of multiple species distribution models could lead to smaller structural uncertainty of fish and fisheries models [@BryndumBuchholz:2019kh]. **Another important source of uncertainty in this study is the utilization of political boundaries (EEZ) to define stocks, rather than biologically-defined populations. While this method might define some stocks that do not necessarily align with biologically-defined sub-populations within an EEZ, in many EEZs, fisheries are often managed at the species level [@MAP:2017uf] and sub-populations are potentially interconnected [@Ramesh:2019va], thus, providing additional ecological ground for our analysis [@Dunn:2019ca]. Moreover, not limiting the species to, at least, an EEZ could result in an overestimation of our results (NOTE: See Ovando *et al.* https://www.pnas.org/content/118/23/e2025958118).** Reproducing our analysis regionally, where spatially explicit stock data is available would allow to generate better constrained results and potentially identify different types of shared stock shifts at meta-population level [@Link:2010ei; @Archambault:2016ie]. Addressing these uncertainties systematically can serve as a roadmap for future studies to provide additional information to inform policy towards sustainable and equitable international fisheries management under climate change.

**NOTE: OPTION B, From a decision-making point of view, these uncertainties hinder the management of natural resources forcing managers to taking action under an uncertain future. Rather than wait for more certain methods to be developed (in some cases is unlikely that we could develop them in the next decade or so, Forlicher *et al.* 2016), manager could build scenario planning into fisheries management [@NOAA_scen_plan_2020]. This will allow for managers to integrate the uncertainty into planning for managing fisheries. For example, if stocks are expected to shift their distribution by 2036 $\pm$ 28 years (Fig. 2), managers could prepare for a scenario where stocks are currently changing and another where they will shift in 20 years. Scenario planning can also be used to estimate potential looses in revenue when adaptation to climate change is not developed cooperatively [@Sumaila:2020tp; @Miller:2013iv].**
